---
title: "Iris Clustering Analysis"
author: "Roshan Adhikari"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
# Ensure ggplot2 is available
if(!require(ggplot2)) install.packages("ggplot2", repos="http://cran.us.r-project.org")
library(ggplot2)
```

## Introduction
This report performs unsupervised K-means clustering on the Iris dataset to identify natural groupings of plants based on their sepal and petal measurements, as outlined in the PRD.

## Data Loading and Validation
First, we load the data and validate the features.

```{r load-data}
file_path <- "data/iris.data"
column_names <- c("sepal_length", "sepal_width", "petal_length", "petal_width", "class")

if (file.exists(file_path)) {
  df <- read.csv(file_path, header = FALSE, col.names = column_names)
  iris_features <- df[, 1:4]
  print("Data loaded successfully.")
} else {
  stop("Data file not found.")
}
```

## Determining Optimal Clusters: Elbow Method
To find the optimal number of clusters ($k$), we evaluate the Total Within-Cluster Sum of Squares (WSS) across different cluster counts.

```{r elbow-plot, fig.height=4, fig.width=6}
max_k <- 10
wss <- numeric(max_k)

for (k in 1:max_k) {
  set.seed(42)
  km <- kmeans(iris_features, centers = k, nstart = 20)
  wss[k] <- km$tot.withinss
}

elbow_df <- data.frame(k = 1:max_k, wss = wss)
ggplot(elbow_df, aes(x = k, y = wss)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue", size = 3) +
  labs(
    title = "Elbow Method for Optimal k",
    x = "Number of Clusters (k)",
    y = "Total Within-Cluster Sum of Squares (WSS)"
  ) +
  theme_minimal() +
  scale_x_continuous(breaks = 1:max_k)
```

## K-means Clustering Results
Based on the elbow plot, the reduction in WSS slows down significantly after **k = 3**, indicating it's an appropriate choice for our number of clusters.

```{r kmeans, fig.height=4, fig.width=6}
k <- 3
set.seed(42)
iris_clusters <- kmeans(iris_features, centers = k, nstart = 20)
df$cluster <- as.factor(iris_clusters$cluster)

# Evaluate against actual labels using a Confusion Matrix
table(Cluster = df$cluster, Actual = df$class)

# Visualization
ggplot(df, aes(x = petal_length, y = petal_width, 
               color = cluster, shape = class)) +
  geom_point(size = 3, alpha = 0.8) +
  labs(
    title = paste("Iris Clustering (K-means, k =", k, ")"),
    subtitle = "Color = Cluster, Shape = Actual Species",
    x = "Petal Length (cm)",
    y = "Petal Width (cm)"
  ) +
  theme_minimal()
```
